---
title: "Crime Rates"
---

Follow along with the demo to learn how to calculate crime rates using R and the Tidyverse. We'll also introduce a new library, janitor, which helps make column names easier to work with. You should get started by loading the tidyverse in a new R block, then install and load janitor.

### Task 1: Load the tidyverse and install and load janitor
```{r}
library(tidyverse)
library(janitor)
```



### Task 2: Load the crime dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_annual_crime.csv

```{r}
md_annual_crimes <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_annual_crime.csv") |> clean_names()
```

### Task 3: Use janitor to clean up the column names

### Task 4: Calculate the total crime rate for each year (picking our unit!)
```{r}
md_annual_crimes <- md_annual_crimes |>
  mutate(crime_per_10k = round((grand_total/population)*10000,0))

md_annual_crimes
```


### Task 5: Calculate the violent and property crime rates for each year (same unit)
```{r}
md_annual_crimes <- md_annual_crimes |>
  mutate(violent_per_10k = round((violent_crime_total/population)*10000,0),
         property_per_10k = round((property_crime_total/population)*10000,0))

md_annual_crimes 
```

```{r}
quantile(data, probs = c(0.25, 0.5, 0.75))
print(quartiles)
```

