---
title: "Crime Rates"
---

Follow along with the demo to learn how to calculate crime rates using R and the Tidyverse. We'll also introduce a new library, janitor, which helps make column names easier to work with. You should get started by loading the tidyverse in a new R block, then install and load janitor.

### Task 1: Load the tidyverse and install and load janitor
```{r}
library(tidyverse)
library(janitor)
```



### Task 2: Load the crime dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_annual_crime.csv

```{r}
md_annual_crimes <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_annual_crime.csv") |> clean_names()
```

### Task 3: Use janitor to clean up the column names

### Task 4: Calculate the total crime rate for each year (picking our unit!)
```{r}
md_annual_crimes <- md_annual_crimes |>
  mutate(crime_per_10k = round((grand_total/population)*10000,0))

md_annual_crimes
```


### Task 5: Calculate the violent and property crime rates for each year (same unit)
```{r}
md_annual_crimes <- md_annual_crimes |>
  mutate(violent_per_10k = round((violent_crime_total/population)*10000,0),
         property_per_10k = round((property_crime_total/population)*10000,0))

md_annual_crimes 
```


### Task 6: Describe our results

Write a few sentences about what you found in the data. What years had the highest crime rates? What years had the lowest crime rates? Does the pattern match your expectations?

It looks like the late 1970s and early 1980s had the highest overall crime rates, and they've steadily gone down since 1999. The lowest crime rates were from 2015-2020. This was shocking to me, I assumed crime would be worse in the last 10 years than previous decades. That finding is also true for property crime, which makes up the vast majority of crimes in Maryland. Violent crime rates have fluctuated more in the last 20 years, peaking in the 90s. The last decade still has the lowest violent crime rates since the 90s, but it has not steadily declined in the same way property crime has.


When you are done, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.

