---
title: "PG Murders and Heat"
author: "Taylor Nichols"
date: "2025-03-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Remember how we talked about people killing each other more when it's hot outside? Let's explore that relationship using homicide and temperature data from Prince George's County. You'll load the data, perform some basic statistical analysis, and create a scatterplot to explore the relationship. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```


### Task 2: Load the dataset from [https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_murder_temps.csv](https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_murder_temps.csv)

```{r}
pg_murders <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_murder_temps.csv")
```

### Task 3: Use `summarize()` and the `cor()` function to calculate the correlation between murders and temperature, using the pearson method.
```{r}
pg_murders |>
  summarize(cor(x = temperature, y = murders, method = "pearson"))
```

### Task 4: Create a scatterplot of murders and temperature, using `geom_point()`. Your aes() will have temperatures on the x axis, murders on the y axis and add a label for the month (so label=month inside the aes()).

```{r}
ggplot(data = pg_murders, aes(x = temperature, y = murders, label=month)) +
  geom_point()
  
```


### Task 5: Add labels to the scatterplot using geom_text(). Inside geom_text() put the same aes() as geom_point in Task 4, then after the aes() add size = 5.

```{r}
ggplot(data = pg_murders, aes(x = temperature, y = murders, label = month)) +
  geom_point() +
    geom_text(
    aes(x = temperature, y = murders, label = month),
    size = 5
  ) 
```


### Task 6: Describe the results of both the correlation calculation and the scatterplot.
Our correlation calculation shows there is a medium correlation. Looking at the scatterplot, I feel like maybe there is a less of a correlation between temperature and perhaps more of one based on the months. It definitely is true that hot months have higher murder rates but it looks like it's less of a case of murder rates correlating directly with temp -- for example, Jan is the coldest month but Nov (a full 10-15 degrees warmer) had lower murder rates. This to me feels like it might be more closely related to something maybe intertwined with temp -- e.g. people being out of school? People spending more time outside? there are a lot of different factors besides temp that I feel like probably come into play here.

When you are done, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.