---
title: "Sampling"
---

Follow along with the demo to learn how to sample data from a population using R and the Tidyverse. You should get started by loading the tidyverse in a new R block.

### Task 1: Load the tidyverse
```{r}
library(tidyverse)
```

### Task 2: Load 1,000 rows of data with the ages of people from 18-80
```{r}
set.seed(42) #what does this do again?

population <- tibble(
  age = sample(18:80, 1000, replace = TRUE)
)
```


### Task 3: Sample 100 rows of data from the dataset
```{r}
sample_100 <- population |>
  sample_n(100)

sample_50 <- population |>
  sample_n(50)

sample_10 <- population |>
  sample_n(10)
```


### Task 4: Calculate the mean age and standard deviation of the sample
```{r}
sample_100 |> summarise(mean = mean(age), sd = sd(age))
sample_50 |> summarise(mean = mean(age), sd = sd(age))
sample_10 |> summarise(mean = mean(age), sd = sd(age))
```


### Task 5: Calculate the mean age and standard deviation of the population
```{r}
population |> summarise(mean = mean(age), sd = sd(age))
```

### Task 6: Compare the sample and population statistics

Compare the means and standard deviations. Are they similar? Go back and re-run the sampling and statistics calculations with a smaller sample size. How do the results change?

They're pretty close for the sample_100, but divert the smaller you make your sample size. Actually what is interesting is the first time I ran my code for sample sizes of 10 and 50, the 50 sample size was pretty far off but the 10 was pretty close to the numbers for the sample size of 100, but I think that was a fluke. It seems like a sample size of 10 percent is decently accurate, and going smaller could be sketchy.



When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
