---
title: "UMD Fees"
---

Follow along with the demo to learn how to calculate percentage change using R and the Tidyverse. You should get started by loading the tidyverse in a new R block.


### Task 1: Load the tidyverse
```{r}
library(tidyverse)
library(janitor)
```

### Task 2: Load the UMD Fees dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/umd_fees.csv

```{r}
umd_fees <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/umd_fees.csv") |> clean_names()
```


### Task 3: Calculate the percentage change in each category between Fall 2024 and Fall 2021 using `mutate()`
```{r}
#pct change formula: (new-old)/old
umd_fees_pct <- umd_fees |> mutate(pct_change = (`Fall 2024`-`Fall 2021`)/`Fall 2021`*100)
```


### Task 4: Describe our results

Write a few sentences about what you found in the data. What's the story here?
Overall, student fees are rising incrementally, with rec services, the shuttle bus and the counseling center accounting for more than half of that increased cost.

I would focus on the shuttle bus increases, which saw the highest dollar amount increase - why are costs rising so much? What is driving that increase? Have services seen a comparable increase? Does the addition of the purple line stand to reduce those costs? How does sustainability play into that? 


When you are done, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
