---
title: "Diversity Index for Maryland Cities"
author: "Taylor Nichols"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
options(scipen = 999)
```

## Overview

In this exercise, we will calculate the **USA TODAY Diversity Index** for Maryland cities with a population of at least 10,000 using 2020 Census race and ethnicity data. The Diversity Index reflects the probability that two randomly selected individuals from a city are of different races or ethnicities.

A score of 0 indicates no diversity (everyone is the same race/ethnicity), while a score closer to 100 indicates higher diversity. The **most diverse city** has the highest index, reflecting a balanced mix of several racial and ethnic groups. The **least diverse** city likely has one overwhelmingly dominant racial/ethnic group.

## Task 1: Load and Describe the Data (3 points)

Run the following code to load the data, then examine it and describe what it contains. Based on the raw numbers alone, which cities stand out as diverse to you?

East Riverdale, Adelphi, Glenmont, Lexington Park, Hyattsville city, Colesville..

```{r load-data}
md_cities <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/usat_diversity_index.csv")

# Preview structure
View(md_cities)
```

## Task 2: Calculate Race Percentages

```{r race-percentages}
md_cities <- md_cities |>
  mutate(
    White_pct = White / TotalPop,
    Black_pct = Black / TotalPop,
    Native_pct = Native / TotalPop,
    Asian_pct = Asian / TotalPop,
    NHPI_pct = NHPI / TotalPop
  )
```

## Task 3: Square the Race Percentages

We need to square the race percentages in order to calculate the probability that any two people chosen at random from a given census area are of different races or are Hispanic or not.

```{r square-race}
md_cities <- md_cities |>
  mutate(
    White_pct_sq = White_pct^2,
    Black_pct_sq = Black_pct^2,
    Native_pct_sq = Native_pct^2,
    Asian_pct_sq = Asian_pct^2,
    NHPI_pct_sq = NHPI_pct^2
  )
```

## Task 4: Calculate Race Probabilities (3 points)

The race probability is the sum the squared probabilities for each of the races. It represents the probability that two people randomly chosen from the same city will be of the same race. Which city has the highest race probability, and why do you think that is?

Summerfield has the highest race probability (87% chance) -- not sure why that is, it's heavily populated by black people/families, maybe something to do with gentrification in DC? Maybe people who got pushed out of southwest moved that way?

```{r race-prob}
md_cities <- md_cities |>
  mutate(
    Race_Prob = White_pct_sq + Black_pct_sq + Native_pct_sq + Asian_pct_sq + NHPI_pct_sq
  ) |> 
  arrange(desc(Race_Prob))

View(md_cities)
```

## Task 5: Calculate Ethnicity Percentages and Their Squares (3 points)

Hispanic origin is not a race but an ethnicity or nationality – and a separate Census question. The probability someone is Hispanic or not Hispanic must be calculated separately from race, and Hisp_Prob is the probability that two randomly selected people will be Hispanic or not. That's an important distinction, because unlike race probability, this measures the likelihood of a binary question: that both random people will be Hispanic. What do you notice about the cities with the *lowest* Hisp_Prob scores?

They appear to all be near UMD? Also 50 percent is still pretty high though I feel like. Although now that I'm looking back at this, maybe this is the percent chance that they will NOT be hispanic -- the cities with the highest hisp_prob (97%) has very few hispanic people, so I feel like this might be the case.

```{r ethnicity}
md_cities <- md_cities |>
  mutate(
    HispTotal = HISPANIC + NotHispanic,
    Hisp_pct = HISPANIC / HispTotal,
    NotHisp_pct = NotHispanic / HispTotal,
    Hisp_pct_sq = Hisp_pct^2,
    NotHisp_pct_sq = NotHisp_pct^2,
    Hisp_Prob = Hisp_pct_sq + NotHisp_pct_sq
  ) |> 
  arrange(Hisp_Prob)

View(md_cities)
```

## Task 6: Multiply Race and Ethnicity Probabilities

Here we multiply the two probabilities together to get the probability that any two random people have the SAME race and ethnicity.

```{r combined-prob}
md_cities <- md_cities |>
  mutate(
    RaceEth_Prob = Race_Prob * Hisp_Prob
  )
```

## Task 7: Calculate Diversity Index (3 points)

The final step is to calculate the probability that two people randomly chosen WILL NOT BE OF THE SAME RACE OR ETHNICITY. In other words, the highest index scores will be cities where those two random people will be from different backgrounds. Put simply, more diverse. Which cities are the top 3, and where are they located? Look at the bottom of the list - which city is the least diverse, and where is it?

Most diverse cities are Langley Park, Wheaton and Adelphi, all just outside of DC near-ish UMD/North of Silver Spring.

Least diverse is Lake Shore, outside of Annapolis.

```{r diversity-index}
md_cities <- md_cities |>
  mutate(
    Diversity_Index = (1 - RaceEth_Prob) * 100
  ) |> 
  arrange(desc(Diversity_Index))

View(md_cities)
```

### Pitch a story (8 points)

Given the data that you've produced here, write a headline and lede of the most interesting story from this exercise, and add below that details on how you would report it out: who you'd talk to and the questions you'd ask. When you describe the Diversity Index values, do so in a similar way to how USA Today did; the raw values themselves won't mean much to readers.

*This was my original answer, but I went back and looked at the numbers again and I think there's a mistake in the description of the Hisp_prob above - maybe I'm wrong but I am abandoning this on that assumption that the 50% hisp_prob is actually the highest probability that two people will be hispanic.*

I feel like I'm having a hard time connecting the dots here in terms of where things are located -- I wish I could map these (I know I could but it would take some extra finagling). Based on what we looked at here, what interested me most was that, despite Maryland being a heavily Black state, it's also heavily Hispanic, which wasn't something I knew or guessed about the state as a whole. I think I would ideally be able to look at historical scores to flesh this out.

*new answer*
Suburbs of Northwest DC most diverse areas in Maryland, according to Census data

Cities on the outskirts of Northwest DC are some of the most diverse in the entire state, according to a new analysis of Census data on race and ethnicity. Residents of cities such as Langley Park, Wheaton, Adelphi and Glenmont are the most diverse in Maryland, with at least a 90 percent chance that any two random people in the area will be of different racial or ethnic backgrounds. The diversity of demographics is spread across the board, with high populations of Black, white, Asian and Hispanic residents in each of these areas.

I would look at what kinds of jobs are located in these areas, where people tend to work, housing prices, and I would interview business and labor experts in the state as well as cultural organizations, immigrant/refugee support services, maybe housing experts/real estate people? And I would also talk to public schools and, like USA Today, probably the library. 

I think I would ask questions like how they have seen demographics change in recent years, how affordability and employment has changed, how the housing market has changed, and what resources people are using. I also think I would talk to people who organize different cultural events to find out if this diversity is resulting in sharing of different cultures, etc. or if people tend to stay pretty segregated.
