---
title: "HW6_Spotify_Songs"
---

In this assignment, you will be working with a dataset of songs played on Spotify in 2020, and you'll be writing code to explore the range of danceability scores in the data, using standard deviation and histograms. You'll also write up your findings.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Load the Spotify Songs dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/spotify_2020.csv

```{r}
songs <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/spotify_2020.csv")
```

### Task 3: Calculate the mean and standard deviation of the danceability scores

Write code to calculate the mean and standard deviation of the danceability scores in the Spotify dataset. Display those.

```{r}
songs |>
  summarise(mean(danceability), median(danceability), sd(danceability))
```

### Task 4: Create a histogram of the danceability scores

Write code to create a histogram of the danceability scores in the Spotify dataset, based on the example we did in class.

```{r}
songs |>
  ggplot() +
  geom_histogram(aes(x = danceability)) +
  geom_vline(aes(xintercept = mean(danceability)), color = "red", linetype = "dashed", size = 1) +  
  geom_vline(aes(xintercept = mean(danceability)-sd(danceability)), color = "blue", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(danceability)+sd(danceability)), color = "blue", linetype = "dashed", size = 1)
```

### Task 5: Add a line to the histogram for the mean danceability score


### Task 6: Add two lines to the histogram to show one standard deviation


### Task 7: Describe our results and pitch a story

Write a few sentences about the distribution of danceability scores in the Spotify dataset. Is it a normal distribution, or is it skewed in one direction? What does the mean and one standard distribution tell us about the songs in the dataset? Draft a short story pitch from this data.

This is not a normal distribution, it's skewed to the right (very danceable!). Would love to compare this to other years. The mean of .65 says the majority of songs played during 2020 were more danceable than not. The SD tells us most songs in 2020 were between .5 and .85 ish. 


Story pitch: People were vibing out during COVID -- the majority of songs made in 2020 were highly danceable, as rated by Spotify, with the majority of songs scoring between 50 and 85 for danceability. Songs that are very danceable include Roses (Remix) by SAINt JHN Feat. Future and WHAT'S POPPIN by Jack Harlow. Less danceable songs include meditation music and Alone Again by The Weeknd.

### Task 8: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
