---
title: "Identifying Schools with Biggest Differences in Math Test Scores, Part 2"
author: "taylor nichols"
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
options(scipen = 999)
```

### Introduction

We'll do what we did in class, but with a twist: instead of Montgomery County, Pa., you'll be working with data from Montgomery County, Md. Same basic idea, though: linear regression looking at free and reduced-price meals and math test scores.

Just as in the exercise, you will create a scatterplot to visualize the relationship between the FARMS (Free and Reduced-price Meal System) percentage and the percentage of students who met or exceeded math expectations. You'll then use linear regression to identify the schools that show the biggest differences between their actual and expected math test scores and evaluate the results of your model.

### Loading the Data

First, let's load the data into a data frame using the read_csv() function from the tidyverse library.

```{r}
schools <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/refs/heads/main/montgomery_md.csv")
```
### Task: Create a Scatterplot and Describe the Results (5 points)

First, let's create a scatterplot to visualize the relationship between the FARMS percentage and the percentage of students who met or exceeded math expectations. We'll use ggplot2 to create the plot and add a linear regression line.

```{r}
schools |>
  ggplot(aes(x = `FARMS Pct`, y = `% Met or Exceeded Math Expectations`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "FARMS Percentage", y = "% Met or Exceeded Math Expectations", title = "Scatterplot of Math Expectations vs FARMS Percentage")
```
Based on what you see here, what's the relationship between FARMS percentage and math test scores? Compare this scatterplot to the one you made in the class activity.

Broadly speaking it looks like there is a negative correlation between the number of FARMS students and math scores. Basically, the higher the rate of FARMS students at a school, the worse the school does on math tests.

This looks very different from what we did in class. It's a much stronger correlation with fewer outliers (random schools with 100 percent FARMS students and decently high math scores).


### Task: Linear Regression Model (7 points)

Next, you'll perform a linear regression analysis using the lm() function. We'll set the percentage of students who met or exceeded math expectations as the dependent variable, and the FARMS percentage as the independent variable. In other words, if we know the FARMS percentage for a school, how reliably can we predict its math score?

```{r}
model <- lm(`% Met or Exceeded Math Expectations` ~ `FARMS Pct`, data = schools)

summary(model)
```

Provide the model's output for the three statistics mentioned in the class activity.

The P-value is extremely low - basically zero (.0000000000000000...22), suggesting the relationship HIGHLY statistically significant. Our R Squared value is pretty high (.67) meaning you 67% of the math scores can be explained by the FARMS percentage.

The coefficients:
Intercept is 68.04878 (b), FARMS coefficient (m) is -0.59511 and y would be the FARMS percentage.

Not entirely sure what we need to do to evaluate these, but I think nothing?

Based on what you've learned so far, what does it tell us about the nature of the relationship between FARMS percentage and math scores? Is this model worth exploring further? Why or why not?

It looks like there is a pretty strong relationship between FARMS percentage and math scores, and it is statistically significant. This model is worth exploring further I believe because of these statistics, although I don't know if our model is a GREAT fit becuase I don't know how to read the residuals number. Oh, looks like we do that next.

### Task: Identifying Schools with Biggest Residuals (3 points)

Let's calculate and examine the residuals next. 

```{r}
schools$predicted <- predict(model)

# Calculate the residuals
schools <- schools |> 
  mutate(residual = `% Met or Exceeded Math Expectations` - predicted)

# Sort the data frame by the absolute value of the residuals in descending order
schools <- schools |>
  mutate(abs_residual = abs(residual)) |>
  arrange(desc(abs_residual))

```

Look at the absolute values of the residuals; what are the largest differences between the predicted result and the actual math scores? Pick two schools with the largest residuals. Find the zip codes where they are located and using the CensusReporter.org website, describe the most interesting demographic data from each zip code.

Wayside and Carderock Springs Elementary Schools have the biggest differences - both are much higher than expected. 

Both are in Montgomery County and are high-income areas. Neither serve many low-income students -- they have hardly any FARMS students. Why did the model predict such high scores for these schools?

## Task: Evaluate the model (5 points)

Describe what conclusions you'd draw from this model's results and how it would influence a story pitch about the school board member and math scores.

I think I would say there is good evidence to support a relationship between FARMS students and math scores, but it seems like there may be some other factors at play. I'm honestly not 100 percent how to interpret this -- the residuals seem crazy to me compared to looking at the scatterplot and knowing what I know about the statistical significance and earlier stats we pulled. I feel like this model isn't a good fit but I think there still is good evidence to support the relationship, although other factors are also likely playing a role.

I don't know that I would rely on this analysis alone to do a story about the school board member and math scores - I think we can safely say there is a correlation between the two and a relationship but I think I would need to do a lot more digging to better understand it.
