---
title: "final_project.rmd"
author: "Taylor Nichols"
date: "2025-04-24"
output: html_document
---

Problems I am having so far:
I am not really sure what to do with this data. I know how to compare it head to head but I am not sure what statistical analysis needs to happen here.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(lubridate)
library(janitor)
```

```{r}
eviction_data <- read_csv("eviction_data_ward_geocodio.csv")
df_violations <- read_delim("detailed_landlord_violations.csv", 
                           delim = "\t",  # Try tab as delimiter
                           locale = locale(encoding = "UTF-16LE"),
                           show_col_types = FALSE) |> clean_names() |>
                            mutate(created_date = case_when(str_detect(created_date, "^\\d{1,2}/\\d{1,2}/\\d{4}$") ~ mdy(created_date)))

```
Let's look at our eviction data 
```{r}
eviction_data
```
Our eviction data is pretty messy. It spans from Nov 2023 to June 2025. It includes address, case number, eviction date, and other geocoding info. 

Now let's look at our violations data:
```{r}
df_violations
```

This data spans from march 2018 to march 2025. We also have owner, violation address, unit, cap_id (not sure what this is), floor, location, violation and fine amt.

I think we can assume that addresses with units are likely owned by the same owner (unless they're condos -- maybe this isn't a safe assumption, actually.) Our landlord data is stored with unit as a separate column, so let's make our data format match.

```{r}
eviction_data_clean <- eviction_data |>
  select(case_number, defendant_address, quad, zipcode, eviction_date, full_address, lat, lng, ward, year, month_name)
```

```{r}
# Transform addresses to match between dataframes
eviction_data_clean <- eviction_data_clean %>%
  mutate(
    # Extract clean street address (without unit)
    clean_address = str_replace_all(
      str_extract(defendant_address, "^[^,#]+"),  # Get everything before first comma or #
      "\\s+,\\s*$", ""  # Remove trailing commas
    ),
    
    # Extract unit number (if present)
    clean_unit = case_when(
      str_detect(defendant_address, "#\\S+") ~ 
        str_replace(str_extract(defendant_address, "#\\S+"), "^#", ""),
      str_detect(defendant_address, ",\\s*[A-Z][0-9]") ~ 
        str_extract(defendant_address, "[A-Z][0-9]"),
      TRUE ~ NA_character_
    )
  )

```

The evictions data is still pretty messy - units will need more cleaning, but at least we can play around with base address now.

```{r}
df_violations_clean <- df_violations |>
  select(s_no, owner, violation_address, unit, created_date, location, violation)
```

Let's make a df that shows us how many violations have happened at each street address and how many units there are. 

```{r}
# Count total violations and unique units per address
df_violations_summary <- df_violations_clean |>
  group_by(violation_address) |>
  summarise(
    total_violations = n(),
    unique_violation_units = n_distinct(unit, na.rm = TRUE),
    unit_violation_list = paste(unique(na.omit(unit)), collapse = ", ")
  )
```

Let's see how many violations there are per owner for fun.
```{r}
df_violations_clean |>
  group_by(owner) |>
  summarise(count=n()) |>
  arrange(desc(count))
```
Now, let's join our violations to our evictions data.
tabling this for now because i need to do some more data cleaning to do it.

```{r}
evictions_summary <- eviction_data_clean |>
  group_by(clean_address) |>
  summarise(
    total_evictions = n(),
    unique_eviction_units = n_distinct(clean_unit, na.rm = TRUE),
    eviction_unit_list = toString(na.omit(unique(clean_unit)))
  )
```

```{r}
summary_df <- full_join(df_violations_summary, evictions_summary, ""
```

