---
title: "final_project.rmd"
author: "Taylor Nichols"
date: "2025-04-24"
output: html_document
---

Kinds of tests I want to do: I think it would be useful to see if there is a correlation between evictions and landlord violations, and then do a linear regression. Once I get my home owner data loaded in, I would also like to see what the relationship between each of the three variables is (we can do that with a multiple regression, right? or just scatterplots on both of them?)

What I really want to get at is how much more likely things are.

I also want to look at eviction rates over time -- maybe weekly and monthly? Would it be worthwhile to look at when spikes occurred? probably. I wish i had more data to do this on but alas I do not.

I also need to pull in the census data/overall stats. Is it worth pulling in census tract level data instead? I kind of feel like that might be more illuminating but also more work which I don't really feel like I need right now. I'll ruminate on it.

I also think it would be wortwhile to look at this at the building level to identify eviction hot spots and how that has changed over time but I don't really know if that's a stats question.

Problems I am having so far: Mostly I am just struggling to get the data all cleaned. The eviction data is a mess. I am getting close but I just need to sit down and have some heads down time on it so I can start my analyses. Which I will do this weekend.

Evictions data comes from the DC Office of the Tenant Advocate. The violations data came from the DC Dept of Buildings Violations and Abatement Tool. <https://dataviz1.dc.gov/t/OCTO/views/DOBPublicDashboard/ViolationsAbatementLVT?%3AshowAppBanner=false&%3Adisplay_count=n&%3AshowVizHome=n&%3Aorigin=viz_share_link&%3Aembed=yes&%3Atoolbar=no>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load libraries}
library(tidyverse)
library(lubridate)
library(janitor)
library(ggplot2)
library(readODS)
library(vcd)
```

## This section is cleaning data and getting it ready to analyze

```{r read in data}
eviction_data <- read_csv("eviction_data_ward_geocodio.csv")
df_violations <- read_delim("detailed_landlord_violations.csv", 
                           delim = "\t",  # Try tab as delimiter
                           locale = locale(encoding = "UTF-16LE"),
                           show_col_types = FALSE) |> clean_names() |>
                            mutate(created_date = case_when(str_detect(created_date, "^\\d{1,2}/\\d{1,2}/\\d{4}$") ~ mdy(created_date)))

homeowners_one <- read_ods("ods_one.ods")
homeowners_two <- read_ods("ods_2.ods")
homeowners_three <- read_ods("ods_3.ods")
homeowners_four <- read_ods("ods_4.ods")
census <- read_csv("census_data.csv")
```

```{r}
dc_homeowners <- bind_rows(homeowners_one, homeowners_two, homeowners_three, homeowners_four)
```

Let's look at our eviction data

```{r}
eviction_data
```

Our eviction data is pretty messy. It spans from Nov 2023 to June 2025. It includes address, case number, eviction date, and other geocoding info.

Now let's look at our violations data:

```{r}
df_violations
```

This data spans from march 2018 to march 2025. We also have owner, violation address, unit, cap_id (not sure what this is), floor, location, violation and fine amt.

I think we can assume that addresses with units are likely owned by the same owner (unless they're condos -- maybe this isn't a safe assumption, actually.) Our landlord data is stored with unit as a separate column, so let's make our data format match.

```{r}
eviction_data_clean <- eviction_data |>
  select(case_number, defendant_address, quad, zipcode, eviction_date, full_address, lat, lng, ward, year, month_name)
```

```{r}
#I used Claude to help me write all this cleaning code
eviction_data_clean <- eviction_data_clean %>%
  mutate(
    # First, separate the unit information from the base address
    base_address = case_when(
      # Extract everything before UNIT, APT, etc.
      str_detect(defendant_address, "(?i)\\s+(UNIT|APT|APARTMENT|SUITE|STE)\\s+[A-Z0-9-]+") ~
        str_trim(str_replace(defendant_address, 
                           "(?i)\\s+(UNIT|APT|APARTMENT|SUITE|STE)\\s+[A-Z0-9-]+.*$", "")),
      
      # If no UNIT/APT keyword, extract everything before a comma or #
      str_detect(defendant_address, "[,#]") ~
        str_trim(str_replace(str_extract(defendant_address, "^[^,#]+"), "\\s+$", "")),
      
      # Default: use the whole address if no unit information is found
      TRUE ~ defendant_address
    ),
    
    # Extract unit information
    unit_info = case_when(
      # Match "UNIT 123" format (works for UNIT 202, UNIT 409, UNIT 101, etc.)
      str_detect(defendant_address, "(?i)\\s+UNIT\\s+\\d+") ~
        str_trim(str_extract(defendant_address, "(?i)\\s+UNIT\\s+\\d+")),
      
      # Match "APT 123" format
      str_detect(defendant_address, "(?i)\\s+APT(ARTMENT)?\\s+\\d+") ~
        str_trim(str_extract(defendant_address, "(?i)\\s+APT(ARTMENT)?\\s+\\d+")),
      
      # Match "STE 123" format
      str_detect(defendant_address, "(?i)\\s+S(UI)?TE\\s+\\d+") ~
        str_trim(str_extract(defendant_address, "(?i)\\s+S(UI)?TE\\s+\\d+")),
      
      # Match "#123" format
      str_detect(defendant_address, "#\\s*[A-Z0-9-]+") ~
        str_trim(str_extract(defendant_address, "#\\s*[A-Z0-9-]+")),
      
      # Match comma followed by unit information
      str_detect(defendant_address, ",\\s*(?i)(UNIT|APT|APARTMENT|SUITE|STE)\\s+\\d+") ~
        str_trim(str_extract(defendant_address, "(?i)(UNIT|APT|APARTMENT|SUITE|STE)\\s+\\d+")),
      
      # Default: NA if no unit information is found
      TRUE ~ NA_character_
    ),
    
    # Extract just the unit number (without the UNIT/APT prefix)
    clean_unit = case_when(
      !is.na(unit_info) & str_detect(unit_info, "(?i)(UNIT|APT|APARTMENT|SUITE|STE)\\s+\\d+") ~
        str_trim(str_replace(unit_info, "(?i)(UNIT|APT|APARTMENT|SUITE|STE)\\s+", "")),
      
      !is.na(unit_info) & str_detect(unit_info, "#\\s*\\d+") ~
        str_trim(str_replace(unit_info, "#\\s*", "")),
      
      TRUE ~ NA_character_
    )
  ) %>%
  # Now clean and standardize the base address
  mutate(
    clean_address = str_replace_all(base_address, "[,\\.]+", ""),
    
    # Standardize street suffixes
    clean_address = case_when(
      str_detect(clean_address, "(?i)\\s+STREET(\\s+|$)") ~ 
        str_replace_all(clean_address, "(?i)\\s+STREET(\\s+|$)", " ST "),
      str_detect(clean_address, "(?i)\\s+AVENUE(\\s+|$)") ~ 
        str_replace_all(clean_address, "(?i)\\s+AVENUE(\\s+|$)", " AVE "),
      str_detect(clean_address, "(?i)\\s+CIRCLE(\\s+|$)") ~ 
        str_replace_all(clean_address, "(?i)\\s+CIRCLE(\\s+|$)", " CIR "),
      str_detect(clean_address, "(?i)\\s+BOULEVARD(\\s+|$)") ~ 
        str_replace_all(clean_address, "(?i)\\s+BOULEVARD(\\s+|$)", " BLVD "),
      str_detect(clean_address, "(?i)\\s+COURT(\\s+|$)") ~ 
        str_replace_all(clean_address, "(?i)\\s+COURT(\\s+|$)", " CT "),
      str_detect(clean_address, "(?i)\\s+DRIVE(\\s+|$)") ~ 
        str_replace_all(clean_address, "(?i)\\s+DRIVE(\\s+|$)", " DR "),
      str_detect(clean_address, "(?i)\\s+LANE(\\s+|$)") ~ 
        str_replace_all(clean_address, "(?i)\\s+LANE(\\s+|$)", " LN "),
      str_detect(clean_address, "(?i)\\s+ROAD(\\s+|$)") ~ 
        str_replace_all(clean_address, "(?i)\\s+ROAD(\\s+|$)", " RD "),
      str_detect(clean_address, "(?i)\\s+PLACE(\\s+|$)") ~ 
        str_replace_all(clean_address, "(?i)\\s+PLACE(\\s+|$)", " PL "),
      str_detect(clean_address, "(?i)\\s+TERRACE(\\s+|$)") ~ 
        str_replace_all(clean_address, "(?i)\\s+TERRACE(\\s+|$)", " TER "),
      str_detect(clean_address, "(?i)\\s+HIGHWAY(\\s+|$)") ~ 
        str_replace_all(clean_address, "(?i)\\s+HIGHWAY(\\s+|$)", " HWY "),
      str_detect(clean_address, "(?i)\\s+PARKWAY(\\s+|$)") ~ 
        str_replace_all(clean_address, "(?i)\\s+PARKWAY(\\s+|$)", " PKWY "),
      str_detect(clean_address, "(?i)\\s+WAY(\\s+|$)") ~ 
        str_replace_all(clean_address, "(?i)\\s+WAY(\\s+|$)", " WY "),
        
      # Default: keep the address as is
      TRUE ~ clean_address
    ),
    
    # Clean up any extra spaces
    clean_address = str_trim(str_replace_all(clean_address, "\\s+", " "))
  )

eviction_data_clean <- eviction_data_clean %>%
  mutate(
    # Extract the street suffix pattern
    suffix_pattern = "(?i)\\s+(ST|AVE|BLVD|CIR|CT|DR|LN|RD|PL|TER|HWY|PKWY|WY)\\b",
    
    # Check if there's any remaining text after the street suffix
    has_extra_text = str_detect(clean_address, paste0(suffix_pattern, "\\s+.+")),
    
    # For addresses with extra text after suffix, update columns
    temp_address = ifelse(has_extra_text,
                         str_extract(clean_address, paste0("^.*?", suffix_pattern)),
                         clean_address),
    
    temp_unit_info = ifelse(has_extra_text,
                           str_trim(str_replace(clean_address, paste0("^.*?", suffix_pattern, "\\s+"), "")),
                           NA_character_),
    
    # Update clean_address with just the part up to the suffix
    clean_address = ifelse(has_extra_text, str_trim(temp_address), clean_address),
    
    # Update unit_info with the extra text if unit_info was previously NA
    unit_info = case_when(
      has_extra_text & is.na(unit_info) ~ temp_unit_info,
      has_extra_text & !is.na(unit_info) ~ paste(unit_info, temp_unit_info, sep = " "),
      TRUE ~ unit_info
    ),
    
    # Update clean_unit to extract numbers from the new unit_info
    clean_unit = case_when(
      # Keep existing clean_unit if already set
      !is.na(clean_unit) ~ clean_unit,
      
      # Extract numbers/alphanumeric from remaining text
      !is.na(unit_info) & str_detect(unit_info, "[0-9A-Z]+") ~
        str_trim(str_extract(unit_info, "[0-9A-Z]+")),
      
      # Extract content from patterns like (D#202)
      !is.na(unit_info) & str_detect(unit_info, "\\([^)]*#([0-9A-Z]+)\\)") ~
        str_trim(str_extract(unit_info, "(?<=\\#)[0-9A-Z]+")),
      
      # Extract T1, D2, etc. type patterns
      !is.na(unit_info) & str_detect(unit_info, "[A-Z][0-9]+") ~
        str_trim(str_extract(unit_info, "[A-Z][0-9]+")),
      
      TRUE ~ NA_character_
    )
  ) %>%
  # Clean up the temporary columns
  select(-suffix_pattern, -has_extra_text, -temp_address, -temp_unit_info) |>
  filter(!str_detect(full_address, "\\bnan\\b")) |>
    filter(!str_detect(full_address, "VACANT LOT"))
```

The evictions data is still pretty messy. Units will need more cleaning, but at least we can play around with base address now.

```{r}
df_violations_clean <- df_violations |>
  select(s_no, owner, violation_address, unit, created_date, location, violation)
```

Let's make a df that shows us how many violations have happened at each street address and how many units there are.

```{r}
# Count total violations and unique units per address
df_violations_summary <- df_violations_clean |>
  group_by(violation_address) |>
  summarise(
    total_violations = n(),
    unique_violation_units = n_distinct(unit, na.rm = TRUE),
    unit_violation_list = paste(unique(na.omit(unit)), collapse = ", ")
  )
```

Let's see how many violations there are per owner for fun.

```{r}
df_violations_clean |>
  group_by(owner) |>
  summarise(count=n()) |>
  arrange(desc(count))
```
Now, let's join our violations to our evictions data. tabling this for now because i need to do some more data cleaning to do it.

```{r}
evictions_summary <- eviction_data_clean |>
  # First combine clean_address and quad (only where quad exists)
  mutate(full_address = ifelse(
    !is.na(quad),
    paste(clean_address, quad),
    clean_address
  )) |>
  # Then continue with your grouping and summarizing
  group_by(full_address) |>
  summarise(
    total_evictions = n(),
    unique_eviction_units = n_distinct(clean_unit, na.rm = TRUE),
    eviction_unit_list = toString(na.omit(unique(clean_unit)))
  )

# Rename the address column in violations summary to match your naming convention
df_violations_summary <- df_violations_summary |>
  rename(full_address = violation_address)
```

```{r}
summary_df <- full_join(df_violations_summary, evictions_summary, by = c("full_address")) |>
    mutate(
    total_violations = replace_na(total_violations, 0),
    total_evictions = replace_na(total_evictions, 0))
```

```{r}
number_of_evictions <- summary_df |>
  group_by(total_evictions) |>
  summarise(number_of_buildings= n())

total_violations <- summary_df |>
  group_by(total_violations) |>
  summarise(number_of_buildings=n())
```

# Data analysis section

## Questions to explore:
Question 1: What is the relationship between scheduled evictions and landlord violations? Is there a correlation between the two?
To answer this I will need to compare whether a property has had a scheduled eviction and a landlord violation (yes/no). I will also need to compare the NUMBER of evictions with the number of landlord violations. I think I can scatterplot both of these (?)


Question 2: What is the relationship between the number of landlord citations and home ownership? Are properties owned by LLCS or LPs more likely to have citations and have MORE citations than properties listed under an individual's name? What types of landlords/what landlords are most likely to have citations? Who are the "worst offenders"?

To answer this I will need a dataframe that tells me the number of citations at a property under each owner. I can get the number of citations under each owner from the citations data, but I need to use the dc homeowners data to determine properties with no citations.

My hypothesis: Property owners/LLCs with many properties are more likely to have citations at their properties than smaller landlords.
 
## Tests to run:
Question 1:
- Scatterplot with a best fit line to look at relationship between scheduled evictions and landlord violations
- Chi-square test to see if evictions are more likely to happen at addresses with housing violations
- Correlation - Use aggregate counts by zip code, ward or census tract to see if there's a relationship between number of violations and number of evictions

Regressions:
Does average amount of fine (take the amount for each cap_id) predicts the eviction count or rate for a ward or zip code?
Logistic regression - do specific factors influence evictions?

Saving these here as part of my project but not really stats questions, more supplementary analysis
- Look at this at the building level to identify eviction hot spots and how that has changed over time
- Look at eviction rates over time -- maybe weekly and monthly?


Question 2:
I am not 100 percent sure how to tackle this - my instinct is to make a bar chart of LLC/LP owned units and avg citations per unit, government-owned properties and citations per unit, and single person landlords and citations per unit to compare but I don't know that that really makes sense?

## Question 1 tests

Scatterplot with a best fit line to look at relationship between scheduled evictions and landlord violations

```{r}
summary_df |>
  ggplot(aes(x = total_violations, y = total_evictions)) +
  geom_point() +
  labs(x = "total_violations", y = "total_evictions", title = "evictions vs violations")
```
This doesn't really seem like it's super easy to analyze in this format given the few big outliers. Let's move on to the Chi Square test for now.


Chi-square test to see if evictions are more likely to happen at addresses with housing violations
```{r}
#had gemini help me with this - i couldn't remember how to make yes/no cols)
categorical_data <- summary_df |>
  mutate(
    has_evictions = ifelse(total_evictions > 0, "Yes", "No"),
    has_violations = ifelse(total_violations > 0, "Yes", "No")
  )
```

Gemini says I don't need to make my own expected frequencies table, and that the chi square test will do it for me. Is this true? I had it help me make one just in case, but wondering if that is a good way to streamline the process. 

```{r}

```


- Correlation - Use aggregate counts by zip code, ward or census tract to see if there's a relationship between number of violations and number of evictions
